Layer dependence as a measure of local dependence

### 2015-04-18 ld9.tex
final changes

\begin{abstract}
A new measure of local dependence called  ``layer dependence" is analysed and illustrated.  Layer dependence measures dependence between two variables at different percentiles in their joint distribution. Layer dependence satisfies coherence properties similar to Spearman's correlation, such as lying between $-1$ and $1$, with $-1$, $0$ and $1$ corresponding to countermonotonicity, independence and comonotonicity, respectively. Spearman's correlation is  a weighted average of layer dependence at different percentiles.  Alternate overall correlation measures are arrived by varying the weights.  Layer dependence allows copulas to be  fitted and tailored to data and expert opinion on the dependence structure.
\end{abstract}

\begin{keyword}
Local dependence; rank dependence; conditional tail expectation; Spearman's correlation; concordance.
\end{keyword}

@@

Dependence between two variables generally varies  with percentile. For example extreme movements in two stock markets are likely to be highly related  whereas minor fluctuations may be relatively independent. Natural catastrophes create significant insurance losses for several classes of business at the same time, while attritional losses between various classes are weakly dependent.

@@

Layer dependence provides a more appropriate and accurate measure of local dependence compared to existing measures. Higher dispersion from various points of the $45^\circ$ line reduces layer dependence and vice versa. Calculating layer dependence at the first instance from data or parametric copulas extracts essential and interpretable information -- the dependence structure. For a parametric copula, the implication of its parametric form and parameters on the dependence structure is not always apparent. Similar problems apply when past data is scarce.

Layer dependence offers an alternative approach to copula modeling. First compute layer dependence values from past data, and apply parametric smoothing. Further adjust, if necessary, to incorporate expert opinion. A copula is then fitted to refined layer dependence values. The fitted copula overcomes the inflexibility of parametric copulas to closely capture the dependence structure in past data, whilst avoiding uncertainties of empirical copulas at the other extreme.

Remaining sections are  as follows. Section \aref{sintroduction} defines and analyzes layer dependence. Section \aref{sldcurve} demonstrates the appropriateness of layer dependence as a local dependence measure using several copulas. Section \aref{sdecompose} explains the behaviour of layer dependence by decomposing it into a negative function of discordance and dispersion. Section \aref{scoherence} describes coherence properties of layer dependence. Links to existing literature are highlighted in  \sref{sliterature}. Further properties of layer dependence are described in  \sref{sproperties}. Section \aref{sfitting} applies layer dependence to copula modeling, and uses historical stock returns as an illustration. Section \aref{saltoverall} discuss alternative coherent measures of overall dependence based on weighted averages of layer dependence. Section \aref{sconclusion} concludes.

@@

A familiar construct in the study of bivariate dependence is Spearman's correlation  \citep{embrechts2002correlation} which is the ordinary correlation between the  ranks of the two variables.   Spearman's correlation $\rho$ measures the correlation between the ranks of two variables and is  one measure in the arsenal for assessing relationships on the percentile or copula scale.   Rank dependence is not distorted by marginal distributions unlike for example Pearson's correlation \citep{mcneil2005qrm} which measures the degree of linear relationship in the original scale. Spearman's $\rho$ can  be applied to estimate copula parameters using the method of moments (\cite{kojadinovic2010comparison}, \cite{bouye2000copulas}). Spearman's correlation suffers from shortcomings and, as an aggregate measure, may be inappropriate for assessing  local dependence which may vary across the joint distribution including the tails.

A familiar construct in reinsurance is a  loss layer.  For example the 95\%-96\% loss layer of a loss $x$ is that portion of loss $x$ between the 95th and 96th percentile
$$
\min\{(x-x_{0.95})^+,x_{0.96}-x_{0.95}\}\ ,
$$
where $()^+$ indicates the positive part of the expression inside the brackets.  Layers arise in the contex of  excess--of--loss reinsurance \citep{wang1995insurance} and debt tranches \citep{mandel2012role}.

A reinsurer  may be interested in the correlation between a  loss layer from an earthquake disaster and say  earthquake severity as measured on  the Richter scale.   The correlation is an imperfect measure as it is scale dependent and ignores the influence of other variables.   Nevertheless it does provide preliminary guidance.

In this paper we combine the  ideas of rank correlation and correlation between one variable and the loss layer of another.   Thus we consider variables on the percentile scale $u$ and $v$, and consider for example the correlation between say the 95\%-96\% loss layer of $u$ and the percentile $v$.   Since variables are on the  percentile scale, the percentile loss layer reduces to
\begin{equation}\label{eg}
\min\{(u-u_{0.95})^+,u_{0.96}-u_{0.95}\} = \min\{(u-\alpha)^+,\de\alpha\}\ , 
\end{equation}
where in this example $\alpha=0.95$ and $\de \alpha=0.01$.

\subsection{Layers on the percentile scale}
The last expression in \eref{eg} can be approximated by, using  the familiar infinitesimal notation,
\begin{equation}\label{ld}
I_\alpha(u)\de \alpha=\de (u-\alpha)^+  \cq I_\alpha(u) \equiv \left\{\begin{array}{cc}0\ , & u\le\alpha\ ,\\ 1\ ,& u>\alpha\ .\end{array}\right.
\end{equation}
The  left hand side expression of \eref{ld} is called the $\alpha$--layer of $u$ at $\alpha$.  (Graph for $(u-\alpha)^+$?)

As often exploited  in reinsurance, any loss or variable can be thought of as a sum of layers:
\begin{equation}\label{decompose}
u=\int_0^u 1\de \alpha= \int_0^1 I_\alpha(u)\de\alpha \ ,
\end{equation}

\subsection{Layer dependence}

Spearman's correlation is the standardized covariance between $v$ and $u$:
$$
\rho = \frac{\cov(v,u)}{\cov(u,u)}
$$
   This numerator equals
\begin{equation}\label{spearman}
\cov(v,u)=\cov\left\{v,\int_0^1I_\alpha(u)\de \alpha \right\}= \int_0^1 \cov\{v,I_\alpha(u)\} \de \alpha \ .
\end{equation}
Write
\begin{equation}\label{ellalpha}
\ell_\alpha \equiv \frac{\cov\{v,I_\alpha(u)\}}{\cov\{u,I_\alpha(u)\}} 
\end{equation}
(*** Insert something here about the insurance meaning of $\ell_\alpha$.)

then 
\begin{equation}\label{definition}
\rho = \int_0^1 \ell_\alpha\cov\{u,I_\alpha(u)\} \de \alpha =  \int_0^1 \ell_\alpha 6\alpha(1-\alpha) \de \alpha \ .
\end{equation}
where the last relation is proved in Appendix xx.

Hence Spearman's $\rho$ is a weighted average of $\ell_\alpha$ across $0\le\alpha\le 1$. The weights are $w_\alpha=6\alpha(1-\alpha)$ and integrate to $1$. Spearman's $\rho$ measures overall dependence between $u$ and $v$ while  $\ell_\alpha$ decomposes the overall $\rho$ into local or layer contributions. Note $w_\alpha$ has minimum $0$ at $\alpha=0$ and $1$, and increases symmetrically to maximum at $\alpha=0.5$. Hence Spearman's $\rho$ places most emphasis on the middle of the distribution which may be undesirable in finance or insurance where tail dependence is critical. Modifying the weights $w_\alpha$ leads to alternate measures of overall dependence further explored in  \sref{saltoverall}.

Layer dependence is a summary measure of a copula.    As any summary measure it can mislead.   However it is not as misleading as for example $\rho$.    Say something about it being a nonparametric summary unlike parameters in a Clayton or Frank copula.  
Mention something about its behaviour for eg Gaussian.

\subsection{Properties of layer--dependence}

Note $\ell_\alpha=1$ if $u=v$,  $\ell_\alpha=-1$ if $u=1-v$ and $u$, $v$ independent implies $\ell_\alpha=0$.   Further $-1\leq \ell_\alpha \leq 1$.   Note also $\ell_\alpha$ is ratio of correlations.

Expanding the covariances  in \eref{definition} and manipulating yields
\begin{equation}\label{gapexp}
\ell_\alpha = \frac{\E(v|u>\alpha)-\E(v|u\leq \alpha)}{\E(u|u>\alpha)-\E(u|u\leq \alpha)}
=2 \left\{\E(v|u>\alpha)-\E(v|u\leq \alpha)\right\}\;,
\end{equation}
where $\E$ calculates expectations. The middle expression in \eref{gapexp} is the expected change in $v$ relative to the expected change in $u$ when $u$ crosses $\alpha$. The latter is $0.5$ for all $\alpha$, yielding the final expression in \eref{gapexp}. Hence large $\ell_\alpha$ implies $v$ is sensitive to movements in $u$ across $\alpha$, indicating strong dependence between $v$ and $u$ at $\alpha$. When $\ell_\alpha=0$, $v$ is unchanged on average when $u$ crosses $\alpha$, and $u$ and $v$ are independent at $\alpha$.

@@

The nine panels in \fref{fillustration} display $(u,v)$ scatterplots of exchangeable copulas $C(u,v)=C(v,u)$, and their layer dependence curves $\ell_\alpha$ for all $0\le\alpha\le 1$. $\ell_\alpha$ and $\alpha$ are plotted on the same axes as $v$ and $u$, respectively. Each copula has Spearman's $\rho=0.6$. \fref{fillustration}  emphasises that copulas with the same overall dependence can a variety of  local dependence structures and that local dependence is  captured with the  layer dependence measure $\ell_\alpha$.

Each $\ell_\alpha$ curve reflects the dependence structure between $u$ and $v$. Given $\alpha$, $\ell_\alpha$ is larger if points are more clustered around $(\alpha,\alpha)$ and vice versa, as formalised in  \sref{sdecompose}. In addition, $\ell_\alpha$ increases to $1$ in the tails if points converge to the $45^\circ$ degree line indicating perfect dependence.

\begin{figure}
  \begin{center}
    \begin{tabular}{ccc}
      \resizebox{40mm}{!}{\includegraphics{normal.eps}}
            \resizebox{40mm}{!}{\includegraphics{student.eps}}
      \resizebox{40mm}{!}{\includegraphics{structural2.eps}} \\
      \resizebox{40mm}{!}{\includegraphics{gumbel.eps}}
            \resizebox{40mm}{!}{\includegraphics{structural1.eps}}
      \resizebox{40mm}{!}{\includegraphics{clayton.eps}} \\
            \resizebox{40mm}{!}{\includegraphics{frank.eps}}
            \resizebox{40mm}{!}{\includegraphics{structural3.eps}}
      \resizebox{40mm}{!}{\includegraphics{structural4.eps}} \\
    \end{tabular}
    \caption{Copulas all  with the same  $\rho=0.6$ but different layer dependence curves $\ell_\alpha$ over $0\leq\alpha\le 1$ (red curves). The top 3 copulas display relatively strong tail dependence in both lower and upper tail.  The first two copulas in the middle three have relatively strong upper tail dependence and while the third has a high degree of lower tail dependence.   The bottom three copulas have relatively high local dependence in the middle of the distribution and less correlation in the tails.  Overall the panels illustrate how an overall given level of $\rho$ can mask a  range of local dependence: and that level dependence is an appropriate tool for assessing local dependence.}
    \label{fillustration}
  \end{center}
\end{figure}

@@

If a copula is symmetric in $u$ and $v$, then $\ell_\alpha$ is the same if $u$ and $v$ are exchanged.   Further, in this exchangeable case,
 layer dependence $\ell_\alpha$ is closely connected to measures the lack of discordance and dispersion.  In particular
\begin{equation}\label{decomposition}
\ell_\alpha = 1-2(1+\gamma_\alpha) \delta_\alpha \;,
\end{equation}
where
$$
\gamma_\alpha \equiv \cor\{(u\leq\alpha),(v>\alpha)\} =\cor\{(u>\alpha),(v\leq\alpha)\}\;,
$$
$$
\delta_\alpha \equiv\E\left\{(|u-v|)|(u-\alpha)(v-\alpha)<0\right\}  \; .
$$
A proof of \eref{decomposition} is  below.  The correlation $-1\leq\gamma_\alpha\leq 1$ measures the tendency for $(u,v)$ to be discordant at $\alpha$: opposite signs on $u-\alpha$ and $v-\alpha$.  The expectation $0\leq \delta_\alpha\leq 1$ measures the dispersion between discordant points $u$ and $v$ at $\alpha$.

The proof of \eref{decomposition} follows from
$$
\gamma_\alpha
=-\frac{\cov\{(u\leq\alpha),(v\leq\alpha)\}}{\var\{(u\leq\alpha)\}}
=\frac{\alpha^2-C(\alpha,\alpha)}{\alpha(1-\alpha)}   \;,
$$
and
$$
\delta_\alpha = 2 \E\left\{(u-v)(u>v)|(u-\alpha)(v-\alpha)<0\right\}
$$
$$
=  \frac{2\E\{(u-v)(u>v)(u>\alpha)(v\leq\alpha)\}}{2\E\{(u>\alpha)(v\leq\alpha)\}}
= \frac{\E\{(u-v)(u>\alpha)(v\leq\alpha)\}}{\alpha-C(\alpha,\alpha)}
$$
$$
= \frac{\E\{(u-v)(u>\alpha)\}-\E\{(u-v)(u>\alpha)(v>\alpha)\}}{\alpha-C(\alpha,\alpha)}
 = \frac{\E\{(u-v)(u>\alpha)\}}{\alpha-C(\alpha,\alpha)} \;.
$$
Substituting the above expressions for $\gamma_\alpha$ and $\delta_\alpha$ into the right hand side of \eref{decomposition} yields the expression for $\ell_\alpha$ in \eref{gapexp}, completing the proof.

Result \eref{decomposition} explains the behaviour of layer dependence curves in \fref{fillustration}. Layer dependence $\ell_\alpha$ is larger if there are fewer discordant pairs at $\alpha$, and discordant pairs at $\alpha$ are closer to the $45^\circ$ degree line. The former indicates smaller $\gamma_\alpha$ and the latter indicates smaller $\delta_\alpha$. Opposite observations apply for small $\ell_\alpha$. If $\ell_\alpha=1$ then $\gamma_\alpha=-1$ or $\delta_\alpha=0$, implying $u$ and $v$ are simultaneously below or above $\alpha$ and $u=v$ for discordant pairs. If $\ell_\alpha=1$  in an interval, then $u=v$  in the same interval.

\fref{finterpretation} illustrates the relationship between layer dependence, discordance and dispersion in \eref{decomposition} using two copulas with $\ell_\alpha=1$ and $\ell_\alpha=0.86$ at $\alpha=0.75$. When $\ell_\alpha=1$, there is no discordance or dispersion at $\alpha$. As $\ell_\alpha$ decreases from $1$, the number of discordant pairs and their dispersion increases.

\begin{figure}
  \begin{center}
    \begin{tabular}{cc}
      \resizebox{60mm}{!}{\includegraphics{perfect.eps}}
      \resizebox{60mm}{!}{\includegraphics{imperf.eps}} \\
    \end{tabular}
    \caption{The left and right panel  show $\ell_{0.75}=1$ and $\ell_{0.75}=0.86$, respectively. In the left panel, $\gamma_{0.75}=-1$ and $\delta_{0.75}=0$. In the right panel, $\gamma_{0.75}=-0.65$ and $\delta_{0.75}=0.21$.  }
    \label{finterpretation}
  \end{center}
\end{figure}

@@

Layer dependence $\ell_\alpha$ satisfies five ``coherence" properties. These properties are  extensions of  properties applying to Spearman's $\rho$.
\begin{itemize}

\item \textbf{Bounds}: Layer dependence lies between $-1$ and $1$: $-1 \le\ell_\alpha \le 1$ for all $\alpha$.
Hence layer dependence is bounded in the same way as  $\rho_S$.

\item \textbf{Perfect dependence}: Constant layer dependence of $-1$ or $1$ are equivalent to countermonotonicity and comonotonicity.
Thus $\ell_\alpha=-1$ for all $\alpha$ if and only if  $v=1-u$ while $\ell_\alpha=1$ for all $\alpha$ if and only if  $v=u$.

\item \textbf{Independence}: If $u$ and $v$ are independent then $\ell_\alpha\equiv 0$.   The converse is not true -- zero layer dependence does not imply independence as shown by the following counterexample. Assume $v=u$ and $v=1-u$ with equal probability. Then $\E(v|u=t)=0.5$ for all $0\leq t\leq 1$ implying $\E(v|u>\alpha)=\E(v|u\leq\alpha)=0.5$. Hence $\ell_\alpha=0$ from \eref{gapexp}. However $u$ and $v$ are not independent.

\item \textbf{Symmetry}: Ranking $v$ in the opposite direction switches the sign of layer dependence. Doing the same to $u$ (the random variable decomposed into layers) switches the sign of layer dependence and flips the layer dependence curve $\ell_\alpha$ about $\alpha=0.5$. Changing the ranking order of both $u$ and $v$ only flips the layer dependence curve.

>>>>

\begin{abstract}
A new measure of local dependence called  ``layer dependence" is proposed and analysed. Layer dependence measures the dependence between two random variables at different percentiles in their joint distribution. Layer dependence satisfies coherence properties similar to Spearman's correlation, such as lying between $-1$ and $1$, with $-1$, $0$ and $1$ corresponding to countermonotonicity, independence and comonotonicity, respectively. Spearman's correlation is  a weighted average of layer dependence across all percentiles.  Alternate overall dependence measures are arrived by varying the weights. Layer dependence is an important input to copula modeling by extracting the dependence structure from past data and incorporating expert opinion if necessary.
\end{abstract}

\begin{keyword}
Local dependence; rank dependence; Spearman's correlation; layers; conditional tail expectation; concordance.
\end{keyword}

@@

Dependence between two random variables generally varies with percentile. For example extreme movements in two stock markets are likely to be highly related  whereas minor fluctuations may be relatively independent. Natural catastrophes create significant insurance losses for several classes of business at the same time, while attritional losses between various classes are weakly dependent.

@@

Layer dependence provides a more appropriate and accurate measure of local dependence compared to existing measures. Higher dispersion between scatter points from the $45^\circ$ line reduces layer dependence and vice versa. For a Gumbel copula exhibiting upper tail dependence, layer dependence starts from a low value and increases to $1$ while the opposite applies to a Clayton copula with lower tail dependence.

Therefore calculating layer dependence at the first instance extracts essential and interpretable information -- the dependence structure. For a parametric copula, the implication of its parametric form and parameters on the dependence structure is not always apparent. Similarly computing layer dependence from data guides the selection and fitting of an appropriate copula.

Remaining sections are as follows. Section \aref{sintroduction} discusses the concepts leading to definition of layer dependence. Section \aref{sldcurve} demonstrates how layer dependence extracts the dependence structure from common copulas. Section \aref{sdecompose} explains the behaviour of layer dependence by decomposing it into a negative function of discordance and dispersion. Section \aref{scoherence} describes coherence properties of layer dependence. Links to existing tail dependence measures are highlighted in  \sref{sliterature}. Further properties of layer dependence are described in  \sref{sproperties}. Section \aref{saltoverall} forms alternate coherent measures of overall dependence apart from Spearman's $\rho$, using weighted averages of layer dependence. Section \aref{sfitting} discusses how layer dependence can be applied to copula modeling. Section \aref{sconclusion} concludes.

@@

This paper combines the concepts of rank correlation and loss layers to form a local dependence measure called layer dependence. The following explains these ideas and the definition of layer dependence sequentially.

A familiar construct in the study of bivariate dependence is Spearman's correlation \citep{embrechts2002correlation}, the linear correlation between ranks of two random variables. Rank dependence avoids distortion from marginal distributions unlike with for example Pearson's correlation \citep{mcneil2005qrm} which measures the degree of linear relationship between random variables in their original values. Spearman's correlation can also be applied to estimate copula parameters using the method of moments (\cite{kojadinovic2010comparison}, \cite{bouye2000copulas}). However Spearman's correlation suffers from shortcomings and, as an aggregate measure, may be inappropriate for assessing local dependence which varies across the joint distribution including the tails.

Another familiar construct, in reinsurance, is a loss layer \citep{wang1995insurance}.  For example the 95\%-96\% layer of a random continuous loss $x$ is the portion of $x$ between its 95th and 96th percentile
$$
\min\{(x-x_{0.95})^+,x_{0.96}-x_{0.95}\}\ ,
$$
where $()^+$ indicates the positive part of the expression inside the brackets and subscripts indicate the percentile. Apart from reinsurance, layers arise in the context of derivative payouts and debt tranches in collaterised debt obligations \citep{mandel2012role}. As the interest is in rank dependence, replace the above quantities with ranks and the 95\%-96\% layer reduces to
\begin{equation}\label{eg}
\min\{(u-u_{0.95})^+,u_{0.96}-u_{0.95}\} = \min\{(u-\alpha)^+,\de\alpha\}\ ,
\end{equation}
where $u$ is the percentile rank transform and in this example $\alpha=0.95$ and $\de \alpha=0.01$, and note $u_\alpha=\alpha$.

\subsection{Percentile layer decomposition}

The final expression in \eref{eg} can be written and approximated by, using the familiar infinitesimal notation,
\begin{equation}\label{ld}
\de \min(u,\alpha) \approx I_\alpha(u)\de \alpha  \cq I_\alpha(u) \equiv \left\{\begin{array}{cc}0\ , & u\le\alpha\ ,\\ 1\ ,& u>\alpha\ .\end{array}\right.
\end{equation}
The left hand side expression of \eref{ld} is called the $\alpha$--layer of $u$. The approximation is an equality in the limit as $\de\alpha \rightarrow 0$. The $\alpha$--layer of $u$ is an infinitesimally small increment $\de\alpha$ if $u$ exceeds $\alpha$ and is zero otherwise. Hence the $\alpha$--layer captures the movement in $u$ at $\alpha$ and ignores movements elsewhere. This interpretation is critical when forming layer dependence discussed below. \fref{flayer} illustrates the $\alpha$--layer of $u$.
\begin{figure}
  \begin{center}
    \begin{tabular}{c}
      \resizebox{100mm}{!}{\includegraphics{layer.eps}}
    \end{tabular}
    \caption{Illustration of $\alpha$-layer of $u$, written as $I_\alpha(u)\de \alpha$. The $\alpha$-layer of $u$ is only sensitive to movements in $u$ at $\alpha$ and ignores other movements.  }
    \label{flayer}
  \end{center}
\end{figure}

As often exploited  in reinsurance, any non-negative random loss or variable can be thought of as a sum of layers. Applying this concept to $u$ yields
\begin{equation}\label{decompose}
u=\int_0^u 1\de \alpha= \int_0^1 I_\alpha(u)\de\alpha    \;.
\end{equation}
Hence $u$ is formed from infinitely many layers, each capturing the movement of $u$ at a particular value.

\subsection{Constructing layer dependence}

Spearman's correlation is the correlation and also the standardized covariance between two percentile rank random variables $v$ and $u$:
$$
\rho \equiv \cor(u,v) = \frac{\cov(v,u)}{\cov(u,u)}
$$
where $\cor$ and $\cov$ indicate covariance and correlation, respectively. In the final expression, the denominator $\cov(u,u)=1/12$ is a constant and a scaling factor ensuring $\rho=1$ if $u=v$ that is $u$ and $v$ are comonotonic. The numerator $\cov(v,u)$ reveals the overall dependence between $v$ and $u$, and is rewritten as, using the decomposition of $u$ in \eref{decompose},
\begin{equation}\label{spearman}
\cov(v,u)=\cov\left\{v,\int_0^1I_\alpha(u)\de \alpha \right\}= \int_0^1 \cov\{v,I_\alpha(u)\} \de \alpha \; .
\end{equation}
Therefore the covariance $\cov(v,u)$ is decomposed into infinitely many covariances $\cov\{v,I_\alpha(u)\}$ for $0\le\alpha\le 1$. Each covariance measures the dependence between $v$ and the $\alpha$--layer of $u$. In a reinsurance setting, this covariance measures dependence between a particular layer of a loss (in percentile rank terms) and another factor. Alternatively the covariance captures dependence between $v$ and movements in $u$ at $\alpha$. Both interpretations point to $\cov\{v,I_\alpha(u)\}$ as a measure of local dependence between $v$ and $u$. Similar to $\rho$, scaling this covariance with the same when $v=u$ leads to the definition of layer dependence
\begin{equation}\label{ellalpha}
\ell_\alpha \equiv \frac{\cov\{v,I_\alpha(u)\}}{\cov\{u,I_\alpha(u)\}}   \;,
\end{equation}
where the denominator 
$$
\cov\{u,I_\alpha(u)\}=\E\{uI_\alpha(u)\}-\E(u)\E\{I_\alpha(u)\}=\int_\alpha^1 u \de u - 0.5\int_\alpha^1 1 \de \alpha
$$
$$
=0.5\alpha(1-\alpha) \;.
$$
where $\E$ calculates expectations.

It is shown that if $\ell_\alpha=1$ for any $\alpha$ then $u$ and $v$ crosses $\alpha$ simultaneously: perfect dependence at $\alpha$. In addition $\ell_\alpha=1$ over $c\le\alpha\le d$ implies $u=v$ over $c\le u,v\le d$. The equality weakens as $\ell_\alpha$ decreases. These results are illustrated and formalised in \sref{sldcurve} and \sref{sdecompose} respectively.

Combining \eref{spearman} and \eref{ellalpha} yields
\begin{equation}\label{wtdavg}
\rho = \int_0^1 \ell_\alpha \frac{\cov\{u,I_\alpha(u)\}}{\cov(u,u)} \de \alpha =  \int_0^1 \ell_\alpha 6\alpha(1-\alpha) \de \alpha \; .
\end{equation}
Hence $\rho$ is a weighted average of $\ell_\alpha$ across $0\le\alpha\le 1$ with weights $w_\alpha=6\alpha(1-\alpha)$ integrating to $1$. This result is consistent with the fact that $\rho$ measures overall dependence whereas $\ell_\alpha$ decomposes $\rho$ into local dependence values. Note $w_\alpha$ has minimum $0$ at $\alpha=0$ and $1$, and increases symmetrically to maximum at $\alpha=0.5$. Hence Spearman's correlation places less emphasis on the tails which may be undesirable in finance or insurance where tail dependence is critical. Modifying the weights $w_\alpha$ leads to alternate measures of overall dependence further explored in \sref{saltoverall}.

Layer dependence summaries the dependence structure of a copula. Layer dependence provides an additional dimension of information compared to Spearman's correlation. As for any summary measure, layer dependence can mislead but is less misleading than $\rho$ and other measures of overall dependence. Layer dependence is also a more meaningful characterisation of a copula compared to the parameters of copula families such as the Clayton or Gumbel \citep{mcneil2005qrm}. Properties of and arguments for using layer dependence are explored below.

This paper assumes exchangeability: $C(u,v)=C(v,u)$ where $C$ is the copula of $(u,v)$. Hence $u$ and $v$ can be interchanged in the above expressions and those remaining in this paper. Archimedean copulas are a well known class of exchangeable copulas \citep{mcneil2005qrm}.

\subsection{Key properties of layer dependence}

Layer dependence $\ell_\alpha$ satisfies the following coherence properties for all $0\le\alpha\le 1$. These properties are shared with Spearman's correlation and are formalised in \sref{scoherence}. Similar to correlation, $0\le\ell_\alpha\le 1$, and $\ell_\alpha=-1$, $0$ and $1$ if $u$ and $v$ are countermonotonic, independent and comonotonic, respectively. In addition $\ell_\alpha$ increases with the correlation order of $(u,v)$. Replacing $u$ or $v$ with their complement leads to straightforward changes in layer dependence.

The following is an alternative expression of $\ell_\alpha$:
\begin{equation}\label{gapexp}
\ell_\alpha = \frac{\E(v|u>\alpha)-\E(v|u\leq \alpha)}{\E(u|u>\alpha)-\E(u|u\leq \alpha)}
=2 \left\{\E(v|u>\alpha)-\E(v|u\leq \alpha)\right\}\;.
\end{equation}
A proof is shown in the appendix. The middle expression in \eref{gapexp} is the expected change in $v$ relative to the expected change in $u$ when $u$ crosses $\alpha$. The latter is $0.5$ for all $\alpha$, yielding the final expression in \eref{gapexp}. Hence large $\ell_\alpha$ implies $v$ is sensitive to movements in $u$ across $\alpha$, indicating strong dependence between $v$ and $u$ at $\alpha$. When $\ell_\alpha=0$, $v$ is unchanged on average when $u$ crosses $\alpha$, hence $u$ and $v$ are not dependent at $\alpha$.

@@

The nine panels in \fref{fillustration} display $(u,v)$ scatterplots of well known exchangeable copulas. Less standard copulas are also shown for illustration. All copulas are calibrated to have equal Spearman's correlation $\rho=0.6$. Layer dependence curves $\ell_\alpha$ for all $0\le\alpha\le 1$ are plotted against $\alpha$ on $(u,v)$ scatterplots to demonstrate the link between the scatter and $\ell_\alpha$.

The scatterplots in \fref{fillustration}  emphasize that copulas with the same overall dependence can exhibit a variety of local dependence structures and these structures are captured by layer dependence curves. Given $\alpha$, $\ell_\alpha$ is larger if scatter points are more clustered around $(\alpha,\alpha)$ and vice versa. In particular $\ell_\alpha$ increases to one in the tails of copulas exhibiting strong tail dependence. Hence $\ell_\alpha$ tracks the clustering of scatter points across the $45\circ$ line. This observation is formalised in \sref{sdecompose}.

\begin{figure}
  \begin{center}
    \begin{tabular}{ccc}
      \resizebox{40mm}{!}{\includegraphics{normal.eps}}
            \resizebox{40mm}{!}{\includegraphics{student.eps}}
      \resizebox{40mm}{!}{\includegraphics{structural2.eps}} \\
      \resizebox{40mm}{!}{\includegraphics{gumbel.eps}}
            \resizebox{40mm}{!}{\includegraphics{structural1.eps}}
      \resizebox{40mm}{!}{\includegraphics{clayton.eps}} \\
            \resizebox{40mm}{!}{\includegraphics{frank.eps}}
            \resizebox{40mm}{!}{\includegraphics{structural3.eps}}
      \resizebox{40mm}{!}{\includegraphics{structural4.eps}} \\
    \end{tabular}
    \caption{Copulas all with the same $\rho=0.6$ but different layer dependence curves $\ell_\alpha$ over $0\leq\alpha\le 1$ (in red). The top 3 copulas display relatively strong tail dependence in both lower and upper tail.  The first two copulas in the middle three have relatively strong upper tail dependence and while the third has a high degree of lower tail dependence. The bottom three copulas have relatively high local dependence in the middle of the distribution and less correlation in the tails. Overall the panels illustrate how a given overall level of $\rho$ can mask a range of dependence structures and that layer dependence is an appropriate tool for assessing local dependence. The top left panel displays a Gaussian copula, followed by a Student's $t$ copula on the right. The leftmost panel in the middle three is a Gumbel copula, and the rightmost is a Clayton copula. The bottom left panel is a Frank copula. Remaining copulas are constructed from a factor model and are included to illustrate how layer dependence captures local dependence.}
    \label{fillustration}
  \end{center}
\end{figure}

@@

As before assume the copula of $(u,v)$ is exchangeable. The following result shows a negative relationship between layer dependence at any point and discordance and dispersion of $(u,v)$ relative to the same point. This result explains the behaviour of layer dependence curves in \fref{fillustration} where $\ell_\alpha$ increases as $(u,v)$ become scattered more closely around $(\alpha,\alpha)$, that is lower discordance and dispersion.

Layer dependence $\ell_\alpha$ can be written as
\begin{equation}\label{decomposition}
\ell_\alpha = 1-2(1+\gamma_\alpha) \delta_\alpha \;,
\end{equation}
where
$$
\gamma_\alpha \equiv \cor\{1-I_\alpha(u),I_\alpha(v)\} =\cor\{I_\alpha(u),1-I_\alpha(v)\}=\frac{\alpha^2-C(\alpha,\alpha)}{\alpha(1-\alpha)}\;,
$$
and
$$
\delta_\alpha \equiv\E\left\{(|u-v|)|(u-\alpha)(v-\alpha)<0\right\}  \; .
$$
A proof of \eref{decomposition} is shown in the appendix. The correlation $-1\leq\gamma_\alpha\leq 1$ measures the tendency for $(u,v)$ to be discordant at $\alpha$: either $u$ or $v$ is above $\alpha$ and the other is below $\alpha$. The expectation $0\leq \delta_\alpha\leq 1$ measures the average dispersion between discordant $u$ and $v$ at $\alpha$, noting $(u-\alpha)(v-\alpha)<0$ is equivalent to $u>\alpha$ and $v\le\alpha$ or $u\le\alpha$ and $v>\alpha$.

\fref{finterpretation} illustrates \eref{decomposition} using two copulas specifically constructed such that, for $\alpha=0.75$, $\ell_\alpha=1$ for one and $\ell_\alpha=0.86$ for the other. For the copula where $\ell_\alpha=1$, there is no discordance or dispersion at $\alpha$. This observation is confirmed by setting $\ell_\alpha=1$ in \eref{decomposition} yielding $\gamma_\alpha=-1$ or $\delta_\alpha=0$. In this case $u$ and $v$ are perfectly dependent at $\alpha$, and are simultaneously below or above $\alpha$. It is also straightforward to extend this result such that if $\ell_\alpha=1$ over $c\le\alpha\le d$ then $u=v$ over $c\le u,v\le d$. As $\ell_\alpha$ decreases from $1$, as for the second copula, the extent of discordance $\gamma_\alpha$ and dispersion $\delta_\alpha$ increases. Again this result can be confirmed from \eref{decomposition} noting $\ell_\alpha$ is negatively related to $\gamma_\alpha$ and $\delta_\alpha$.

\begin{figure}
  \begin{center}
    \begin{tabular}{cc}
      \resizebox{60mm}{!}{\includegraphics{perfect.eps}}
      \resizebox{60mm}{!}{\includegraphics{imperf.eps}} \\
    \end{tabular}
    \caption{The left and right panel show copulas where $\ell_{0.75}=1$ and $\ell_{0.75}=0.86$, respectively. In the left panel, $\gamma_{0.75}=-1$ (no discordant points) and $\delta_{0.75}=0$ (zero dispersion). In the right panel, $\gamma_{0.75}=-0.65$ (some discordant points) and $\delta_{0.75}=0.21$ (some dispersion between discordant points). }
    \label{finterpretation}
  \end{center}
\end{figure}

The result in \eref{decomposition} explains the behaviour of layer dependence curves in \fref{fillustration}. Layer dependence $\ell_\alpha$ is larger if scatter points are more clustered around $(\alpha,\alpha)$, that is, fewer discordant points at $\alpha$, and discordant points at $\alpha$ are closer to the $45^\circ$ degree line. The former indicates smaller $\gamma_\alpha$ and the latter indicates smaller $\delta_\alpha$. Opposite observations apply for small $\ell_\alpha$.

@@

Layer dependence $\ell_\alpha$ satisfies the following five ``coherence" properties. These properties are extensions of properties applying to Spearman's correlation. 
\begin{itemize}

\item \textbf{Bounds}: Layer dependence lies between $-1$ and $1$: $-1 \le\ell_\alpha \le 1$ for all $\alpha$.
Hence layer dependence is bounded in the same way as $\rho$.

\item \textbf{Perfect dependence}: Constant layer dependence of $-1$ or $1$ are equivalent to countermonotonicity and comonotonicity, respectively. Thus $\ell_\alpha=-1$ for all $\alpha$ if and only if  $v=1-u$ while $\ell_\alpha=1$ for all $\alpha$ if and only if  $v=u$.

\item \textbf{Independence}: If $u$ and $v$ are independent then $\ell_\alpha=0$ for all $\alpha$. The converse is not true -- zero layer dependence does not imply independence as shown by the following counterexample. Assume $v=u$ and $v=1-u$ with equal probability. Then $\E(v|u=\alpha)=0.5$ for all $0\leq \alpha\leq 1$ implying $\E(v|u>\alpha)=\E(v|u\leq\alpha)=0.5$. Hence $\ell_\alpha=0$ from \eref{gapexp}. However $u$ and $v$ are not independent.

\item \textbf{Symmetry}: Replacing $v$ with $1-v$ yields layer dependence curve $-\ell_\alpha$. Doing the same to $u$ (the random variable decomposed into layers) yields layer dependence curve $-\ell_{1-\alpha}$ hence a flip is performed about $\alpha=0.5$ in addition to a sign change. Replacing both $u$ and $v$ with their complements yields layer dependence curve $\ell_{1-\alpha}$.
